---
title: 'Comparative Study: R, PostgreSQL, MongoDB'
author: "Daniel Dittenhafer"
date: "Thursday, October 30, 2014"
output: html_document
---

### Data Set Use Case ###

### Advantages / Disadvantages ###

### Recommendation ###

### Appendix ###

#### Data Set Description ####
The data set used in this comparative study was obtained from the U.S. Department of Housing and 
Urban Development (HUD) via their Data Sets Portal at http://www.huduser.org/portal/datasets/fmr.html. 
The specific data set used is the 2015 Fair Market Rents Country Level Data File:

http://www.huduser.org/portal/datasets/fmr/fmr2015f/FY2015F_4050_Final.xls

The data set was downloaded in Microsoft Excel format (XLS) and saved to CSV format to enable
easier access by the various technologies used throughout this comparative study.




```{r}
summary(cars)
```

You can also embed plots, for example:

```{r, echo=FALSE}
plot(cars)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

### Source Code ###
The raw R markdown and other code used in this comparative study can be found 
[on GitHub, in my DataAcqMgmt repository](https://raw.githubusercontent.com/dwdii/DataAcqMgmt/master/Week10).

### References ###

