---
title: 'Comparative Study: R, PostgreSQL, MongoDB'
author: "Daniel Dittenhafer"
date: "Thursday, October 30, 2014"
output: pdf_document
---

### Data Set Use Case ###

The data set used in this comparative study was obtained from the U.S. Department of Housing and 
Urban Development (HUD) contains fair market rent (FMR) data points for counties and 
metropolitan statisitcal areas through-out the United States of America. The FMR data points
are required to be published by as part of Section 8 of the United States Housing Act of 1937.

The FMR data is used by programs operating under Section 8 including the Housing Choice Voucher,
the Moderate Rehabilitation, and the project-based voucher programs. External to these programs,
this data set can be used to better understand differences in housing costs around the country,
and if compared to other years, how costs are changing over time.

### Advantages / Disadvantages ###

#### R ####

The R runtime provides several functions for saving and loading data in a R specific format. The `save` function
will persist one or more variables to a specified file in either ASCII or binary format, and likewise
the `load` function will deserialize the data from a given file into the R runtime's environment.

These functions are quite useful for preserving work from day to day, or sharing small data sets
between colleagues. Mid-sized data sets could be serialized using these functions 
as well, wherein the compress option might be a significant benefit. 

Beyond the use cases described above, the RData format is unlikely to be useful 
for larger data sets (8+ GiB). The R runtime wants to place any data loaded from
RData files into the environment's memory space. Depending on your operating system
some of this may be swapped to disk, but nonetheless working memory is now limited
due to the presence of the loaded data.

Simultaneous access by multiple users could also be a problem. Although once a data set
is loaded from a given file, another user could also load the same data from the file,
for example if the file is shared on a network drive, sharing changes presents a problem.
The entire data set whould need to be serialized back to the file and notifications broadcast
to collegues of then need to reload the file.

#### PostgreSQL ####

[PostgreSQL](http://www.postgresql.org/about/) is an open source relational database management
system which stores data in databases and tables hosted through a separate data service typically 
on a separate server.

PostgreSQL, and relational database systems in general, are well suited for persisting data sets ranging 
from kilobytes through gigabytes to terabytes. According to the PostgreSQL website, a given database in
not limited in size, and a given table with a database can be as large as 32TB. 

#### MongoDB ####

### Recommendation ###

### Appendix ###

#### Data Set Description ####
 
The specific data set used is the 2015 Fair Market Rents Country Level Data File:

http://www.huduser.org/portal/datasets/fmr/fmr2015f/FY2015F_4050_Final.xls

The data set was downloaded in Microsoft Excel format (XLS) and saved to CSV format to enable
easier access by the various technologies used throughout this study.

----------------------------------------------------------------------------------------------------
Data Set         Number of              Attribute              Number of 
Characteristics  Observations Area      Characteristics        Attributes Missing Values?
--------------- ------------- --------- ---------------------- ---------- ---------------
Multivariate    4,769         Business  Categorical, Integer   18         Yes    
                                          
--------------------------------------- ---------------------- ---------- ---------------

#### Attribute Information ####

The following table lists the data types and a brief description of the fields contained 
in the data set used in this comparative study.

-------------------------------------------------------------------------------
Field                Data Type     Description
-------------------  ------------- --------------------------------------------
fips2000             integer       2000 Census 10-digit state, county, metro code

fips2010             integer       2010 Census 10-digit state, county, metro code

fmr2                 integer       2 bedroom fair market rent

fmr0                 integer       0 bedroom (studio) fair market rent

fmr1                 integer       1 bedroom fair market rent

fmr3                 integer       3 bedroom fair market rent

fmr4                 integer       4 bedroom fair market rent

county               integer       County number

State                integer       State number

CouSub               integer       County Sub number

countyname           string        County Name

Metro_code           string        METROxxxxxxMxxxxx for metropolitan areas and 
                                   NCNTYxxxxxNxxxxx for nonmetropolitan counties

Areaname             string        Area name
 
county_town_name     string        ???

pop2010              integer       Population of locality from 2010 Census

state_alpha          string        2 character abbreviation for State

fmr_type             integer       Percentile of FMR  

metro                binary        1 = Metro, 0 = non-metro county

---------------------------------------------------------------------

#### R Code ####
The following R code demonstrates loading the data set from CSV format, manipulating
to ensure proper data typing

```{r, echo=FALSE}
require(ggplot2, quietly=TRUE)
projRoot <- "C:/Code/R/DataAcqMgmt"

# Disable warnings
options(warn=-1)  
```

```{r}
# Load the data into a data.frame
csv_file <- file.path(projRoot, "Week10", "Data", "FY2015F_4050_Final.csv")
csv <- read.table(csv_file, header=TRUE, sep=",", 
                  fill=TRUE, stringsAsFactors=FALSE, encoding="UTF-8")

# Data types and scaling
dfFmrData <- data.frame(csv)
dfFmrData$fmr0 <- as.numeric(csv$fmr0)
dfFmrData$fmr1 <- as.numeric(csv$fmr1)
dfFmrData$fmr2 <- as.numeric(csv$fmr2)
dfFmrData$fmr3 <- as.numeric(csv$fmr3)
dfFmrData$fmr4 <- as.numeric(csv$fmr4)
dfFmrData$metro <- as.numeric(csv$metro)
dfFmrData$pop2010 <- as.numeric(csv$pop2010) / 1000000

# Save to RData format
savePath <- file.path(projRoot, "Week10", "Data", "FMR_FY2015_Final.RData")
save(dfFmrData, file=savePath)

# Load from RData format
load(savePath)

```

```{r, echo=FALSE}
g5 <- ggplot(data=dfFmrData, aes(x=pop2010))
g5 <- g5 + geom_point(aes(y=fmr0, colour="Studio"))
g5 <- g5 + geom_point(aes(y=fmr2, colour="2 Bedroom"))
g5 <- g5 + geom_point(aes(y=fmr3, colour="3 Bedroom"))
g5 <- g5 + scale_fill_hue(l=40)
g5 <- g5 + theme(axis.text.x = element_text(angle=30, vjust=1))
g5 <- g5 + guides(colour = guide_legend("Legend"))
g5 <- g5 + labs(title="Fair Market Rent vs. 2010 Census",
                x="2010 Census Population (Millions)",
                y="Fair Market Rent (USD)")
g5

```

```{r, echo=FALSE}
summary(dfFmrData)
```

### Source Code ###

The raw R markdown and other code used in this comparative study can be found 
[on GitHub, in my DataAcqMgmt repository](https://raw.githubusercontent.com/dwdii/DataAcqMgmt/master/Week10).

### References ###

"About." PostgreSQL:. N.p., n.d. Web. 02 Nov. 2014. <http://www.postgresql.org/about/>.

"Final Fair Market Rents for the Housing Choice Voucher Program and Moderate Rehabilitation Single Room Occupancy Program Fiscal Year 2015, Docket No. FR-5807-N-03; Notice of Final Fiscal Year (FY) 2015 Fair Market Rents (FMRs)," 70 Federal Register 192 (3 October 2014), pp 59786-59792.

"R: Save R Objects." R: Save R Objects. R Project, n.d. Web. 02 Nov. 2014. <http://stat.ethz.ch/R-manual/R-patched/library/base/html/save.html>.

"What Are the Disadvantages of Using .Rdata Files Compared to HDF5 or NetCDF?" StackOverflow.com. Stack Exchange Inc., 25 Oct. 2011. Web. 02 Nov. 2014. <http://stackoverflow.com/a/7890475>.