---
title: 'Comparative Study: R, PostgreSQL, MongoDB'
author: "Daniel Dittenhafer"
date: "Thursday, October 30, 2014"
output: pdf_document
---

### Data Set Use Case ###

The data set used in this comparative study was obtained from the U.S. Department of Housing and 
Urban Development (HUD) contains fair market rent (FMR) data points for counties and 
metropolitan statisitcal areas through-out the United States of America. The FMR data points
are required to be published by as part of Section 8 of the United States Housing Act of 1937.

The FMR data is used by programs operating under Section 8 including the Housing Choice Voucher,
the Moderate Rehabilitation, and the project-based voucher programs. External to these programs,
this data set can be used to better understand differences in housing costs around the country,
and if compared to other years, how costs are changing over time.

### Advantages / Disadvantages ###

### Recommendation ###

### Appendix ###

#### Data Set Description ####
 
The specific data set used is the 2015 Fair Market Rents Country Level Data File:

http://www.huduser.org/portal/datasets/fmr/fmr2015f/FY2015F_4050_Final.xls

The data set was downloaded in Microsoft Excel format (XLS) and saved to CSV format to enable
easier access by the various technologies used throughout this study.

----------------------------------------------------------------------------------------------------
Data Set         Number of              Attribute              Number of 
Characteristics  Observations Area      Characteristics        Attributes Missing Values?
--------------- ------------- --------- ---------------------- ---------- ---------------
Multivariate    4,769         Business  Categorical, Integer   18         Yes    
                                          
--------------------------------------- ---------------------- ---------- ---------------

#### Attribute Information ####

-------------------------------------------------------------------------------
Field                Data Type     Description
-------------------  ------------- --------------------------------------------
fips2000             integer       2000 Census 10-digit state, county, metro code

fips2010             integer       2010 Census 10-digit state, county, metro code

fmr2                 integer       2 bedroom fair market rent

fmr0                 integer       0 bedroom (studio) fair market rent

fmr1                 integer       1 bedroom fair market rent

fmr3                 integer       3 bedroom fair market rent

fmr4                 integer       4 bedroom fair market rent

county               integer       County number

State                integer       State number

CouSub               integer       County Sub number

countyname           string        County Name

Metro_code           string        METROxxxxxxMxxxxx for metropolitan areas and 
                                   NCNTYxxxxxNxxxxx for nonmetropolitan counties

Areaname             string        Area name
 
county_town_name     string        ???

pop2010              integer       Population of locality from 2010 Census

state_alpha          string        2 character abbreviation for State

fmr_type             integer       Percentile of FMR  

metro                binary        1 = Metro, 0 = non-metro county

---------------------------------------------------------------------

#### R Code ####

```{r, echo=FALSE}
projRoot <- "C:/Code/R/DataAcqMgmt"

```{r}
# Load the data into a data.frame
csv_file <- file.path(projRoot, "Week10", "Data", "FY2015F_4050_Final.csv")
csv <- read.table(csv_file, header=TRUE, sep=",")

summary(csv)
```

You can also embed plots, for example:

```{r, echo=FALSE}
plot(cars)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

### Source Code ###
The raw R markdown and other code used in this comparative study can be found 
[on GitHub, in my DataAcqMgmt repository](https://raw.githubusercontent.com/dwdii/DataAcqMgmt/master/Week10).

### References ###

"Final Fair Market Rents for the Housing Choice Voucher Program and Moderate Rehabilitation Single Room Occupancy Program Fiscal Year 2015, Docket No. FR-5807-N-03; Notice of Final Fiscal Year (FY) 2015 Fair Market Rents (FMRs)," 70 Federal Register 192 (3 October 2014), pp 59786-59792.

