---
title: 'Comparative Study: R, PostgreSQL, MongoDB'
author: "Daniel Dittenhafer"
date: "Thursday, October 30, 2014"
output: html_document
---

### Data Set Use Case ###

### Advantages / Disadvantages ###

### Recommendation ###

### Appendix ###

#### Data Set Description ####
The data set used in this comparative study was obtained from the U.S. Department of Housing and 
Urban Development (HUD) via their Data Sets Portal at http://www.huduser.org/portal/datasets/fmr.html. 
The specific data set used is the 2015 Fair Market Rents Country Level Data File:

http://www.huduser.org/portal/datasets/fmr/fmr2015f/FY2015F_4050_Final.xls

The data set was downloaded in Microsoft Excel format (XLS) and saved to CSV format to enable
easier access by the various technologies used throughout this comparative study.

----------------------------------------------------------------------------------------------------
Data Set         Number of              Attribute              Number of 
Characteristics  Observations Area      Characteristics        Attributes Missing Values?
--------------- ------------- --------- ---------------------- ---------- ---------------
Multivariate    4,769         Business  Categorical, Integer   18         Yes    
                                          
--------------------------------------- ---------------------- ---------- ---------------

#### Attribute Information ####

-------------------------------------------------------------------------------
Field                Data Type     Description
-------------------  ------------- --------------------------------------------
fips2000             integer       2000 Census 10-digit state, county, metro code

fips2010             integer       2010 Census 10-digit state, county, metro code

fmr2                 integer       2 bedroom fair market rent

fmr0                 integer       0 bedroom (studio) fair market rent

fmr1                 integer       1 bedroom fair market rent

fmr3                 integer       3 bedroom fair market rent

fmr4                 integer       4 bedroom fair market rent

county               integer       County number

State                integer       State number

CouSub               integer       County Sub number

countyname           string        County Name

Metro_code           string        METROxxxxxxMxxxxx for metropolitan areas and 
                                   NCNTYxxxxxNxxxxx for nonmetropolitan counties

Areaname             string        Area name
 
county_town_name     string        ???

pop2010              integer       Population of locality from 2010 Census

state_alpha          string        2 character abbreviation for State

fmr_type             integer       Percentile of FMR  

metro                binary        1 = Metro, 0 = non-metro county

---------------------------------------------------------------------

```{r}
summary(cars)
```

You can also embed plots, for example:

```{r, echo=FALSE}
plot(cars)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

### Source Code ###
The raw R markdown and other code used in this comparative study can be found 
[on GitHub, in my DataAcqMgmt repository](https://raw.githubusercontent.com/dwdii/DataAcqMgmt/master/Week10).

### References ###

