---
title: "Which year were the best popular movies made?"
author: "Daniel Dittenhafer"
date: "Tuesday, September 16, 2014"
output: html_document
---
```{r, echo=FALSE}
require(ggplot2)
```

Which year were the best popular movies made? 1987? 2002? 1955? Using R and a 
[dataset provided by Hadley Wickham](http://had.co.nz/data/movies/), I have come
to a conclusion. But first let me describe my process.

The data set, originally from [IMDB.com](http://IMBD.com/), provides a variety 
of data points about each movie going back as far as 1893 (but is current only 
through 2005). Some data points of interest include IMDB user rating, the 
number of rating votes cast, the budget for making the film, genre and MPAA rating.

Given that this data set comes originally from IMDB.com, an internet website, the user ratings are 
most likely biased to those movies for which internet users find compelling in some way. 
With that acknowledged, lets proceed.

My approach is to use the `votes` data point as a proxy for popularity. Hopefully this
is intuitive, but the concept is that if a movie is popular, good or bad, it will 
receive more votes.

The following code illustrates loading the data set and massaging the data to 
tease out the number of movies and votes cast per year. I divide the total votes
per year by the number of movies in the year as a technique to normalize against 
years with more movies and potentially more votes as a result.

```{r}
# Load the movie data set
movies <- read.table("C:/SourceCode/R/DataAcqMgmt/MovieData/movies.tab", sep="\t", header=TRUE, quote="", comment="")

# Group by year and sum votes 
votesByYear <- aggregate(votes ~ year, movies, sum)

# Using plyr::count to determine movies/yr
moviesByYear <- plyr::count(movies, "year")
moviesVotesByYear <- merge(votesByYear, moviesByYear, by="year")

# Divide votes by # movies in the year to account for years with more 
# movies contributing to more total votes
avgVotesPerMovie <- moviesVotesByYear$votes / moviesVotesByYear$freq
moviesVotesByYear <- cbind(moviesVotesByYear, avgVotesPerMovie)

# Sort votes descending and grab the 10 years with most votes
votesByYearSorted <- moviesVotesByYear[order(-moviesVotesByYear$avgVotesPerMovie), ]
mostPopularByVote <- votesByYearSorted[1:10, ]

print(mostPopularByVote)
```

```{r, echo=FALSE}
g1 <- ggplot(data=mostPopularByVote, aes(x=factor(year), y=avgVotesPerMovie))
g1 <- g1 + geom_bar(stat="identity")
g1 <- g1 + scale_x_discrete(labels=as.character(mostPopularByVote$year), limits=as.character(mostPopularByVote$year))
g1 <- g1 + labs(title="Top 10 Average Votes per Movie", x="Year", y="Average Votes / Movie")
g1

```

Now that we have a list of candidate years with popular movies, the `rating` data point 
can be brought in as a proxy for "best". The following R code first calculates the
mean rating per year across all movies and then pulls out the years that were found
to be most popular based on the `votes` metric. Finally, the popular movies
are sorted by rating descending.

```{r}
# Determine average movie rating per year
avgRatings <- aggregate(rating ~ year, movies, mean)

# Extract only the years from our top 10 most popular and sort by avg rating.
popMoviesAvgRating <- subset(avgRatings, year %in% mostPopularByVote$year, year:rating)
popMoviesAvgRating <- popMoviesAvgRating[order(-popMoviesAvgRating$rating), ]
print(popMoviesAvgRating)

```

```{r, echo=FALSE}
g2 <- ggplot(data=popMoviesAvgRating, aes(x=factor(year), y=rating))
g2 <- g2 + geom_bar(stat="identity")
g2 <- g2 + scale_x_discrete(labels=as.character(popMoviesAvgRating$year),
                            limits=as.character(popMoviesAvgRating$year))
g2 <- g2 + labs(title="Top 10 Years for Best Popular Movies", x="Year", y="Average Rating / Year")
g2

```

Based on the analysis described above, the year 2002 appears to be the year when
the best popular movies were made. Using the following R code, I extract the 
specific list of the best popular movies.

```{r}
# Extract movies for the specific year and sort by votes and rating
moviesIn2002 <- subset(movies, year == 2002, c("title", "votes", "rating"))
moviesIn2002Sorted <- moviesIn2002[order(-moviesIn2002$votes, -moviesIn2002$rating),]

# Pull out top 10
top10MoviesIn2002 <- data.frame(moviesIn2002Sorted[1:10,])

```

```{r, result='asis', echo=FALSE}
knitr::kable(top10MoviesIn2002)
```

Links I found useful while working this assignment:

* [StackOverflow: How to Change the Order of a Discrete X Scale in ggplot](http://stackoverflow.com/questions/3253641/how-to-change-the-order-of-a-discrete-x-scale-in-ggplot)

Best,

Daniel

