---
title: "Which year were the best popular movies made?"
author: "Daniel Dittenhafer"
date: "Tuesday, September 16, 2014"
output: html_document
---
```{r, echo=FALSE}
require(ggplot2)
```

Which year were the best popular movies made? 1987? 2002? 1955? Using R and a 
[dataset provided by Hadley Wickham](http://had.co.nz/data/movies/), I have come
to a conclusion. But first let me describe my process.

The dataset, originally from [IMDB.com](http://IMBD.com/), provides a variety 
of data points about each movie going back as far as 1893 (but is current only 
through 2005). Some data points of interest include IMDB user rating, the 
number of rating votes cast, the budget for making the film, genre and MPAA rating.

Given that this dataset comes originally from IMDB.com, an internet website, the user ratings are 
most likely biased to those movies for which internet users find compelling in some way. 

My approach is to use the `votes` data point as a proxy for popularity. Hopefully this
is intuitive, but the concept is that if a movie is popular, good or bad, it will 
receive more votes.

```{r}
# Load the movie data st
movies <- read.table("C:/SourceCode/R/DataAcqMgmt/MovieData/movies.tab", sep="\t", header=TRUE, quote="", comment="")

# Group by year and sum votes and # movies/yr
votesByYear <- aggregate(votes ~ year, movies, sum)
moviesByYear <- aggregate(movies$year, by=list(year=movies$year), FUN=length)
names(moviesByYear) <- c("year", "count")
moviesVotesByYear <- merge(votesByYear, moviesByYear, by="year")

# Divide votes by # movies in the year to account for years with more 
# movies contributing to more total votes
avgVotesPerMovie <- moviesVotesByYear$votes / moviesVotesByYear$count
moviesVotesByYear <- cbind(moviesVotesByYear, avgVotesPerMovie)

# Sort votes descending and grab the 10 years with most votes
votesByYearSorted <- moviesVotesByYear[order(-moviesVotesByYear$avgVotesPerMovie), ]
mostPopularByVote <- votesByYearSorted[1:10, ]

print(mostPopularByVote)
```

```{r, echo=FALSE}
g1 <- ggplot(data=mostPopularByVote, aes(x=factor(year), y=avgVotesPerMovie))
g1 <- g1 + geom_bar(stat="identity")
g1 <- g1 + scale_x_discrete(labels=as.character(mostPopularByVote$year), limits=as.character(mostPopularByVote$year))
g1 <- g1 + labs(title="Top 10 Average Votes per Movie", x="Year", y="Average Votes / Movie")
g1

```

Now that we have a list of candidate years with popular movies, the `rating` data point 
can be brought in as a proxy for "best". 

```{r}



```

You can also embed plots, for example:

```{r, echo=FALSE}
plot(cars)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
