---
title: 'Data Set Profile: Bike Sharing Demand'
author: "Daniel Dittenhafer"
date: "Tuesday, September 23, 2014"
output: html_document
---
```{r, echo=FALSE} 
require(ggplot2, quietly=TRUE)
require(lubridate, quietly=TRUE)
require(DMwR, quietly=TRUE)
projRoot <- "C:/Code/R/DataAcqMgmt"
```
----------------------------------------------------------------------------------------------------
Data Set         Number of              Attribute                        Number of 
Characteristics  Observations Area      Characteristics                  Attributes  Missing Values?
--------------- ------------- --------- -------------------------------- ----------- ---------------
Multivariate    10,886        Business  Categorical, Integer, Date/time  12          No     
--------------------------------------- -------------------------------- ----------- ---------------

### Source ###

[Kaggle.com Competition: Bike Sharing Demand](https://www.kaggle.com/c/bike-sharing-demand)


### Attribute Information ###

--------------------------------------------------------------------
Field      Data Type     Description
---------  ------------- --------------------------------------------
datetime   date & hour   Min: 01/01/2011 00:00;    
                         Max: 12/19/2012 23:00

season     integer       Categorical;
                         1 = spring;
                         2 = summer;
                         3 = fall;
                         4 = winter;
                         
holiday    boolean       1 = a holiday;
                         0 = not a holiday;

workingday boolean       1 = a work day;
                         0 = weekend or holiday;
                         
weather    integer       Categorical;
                         1) Clear, Few clouds, Partly cloudy, Partly cloudy;
                         2) Mist + Cloudy, Mist + Broken clouds, Mist + Few clouds, Mist; 
                         3) Light Snow, Light Rain + Thunderstorm + Scattered clouds, Light Rain + Scattered clouds;
                         4) Heavy Rain + Ice Pallets + Thunderstorm + Mist, Snow + Fog; 
                         
temp       decimal       temperature in Celsius

atemp      decimal       apparent temperature in Celsius 

humidity   integer       relative humidity percentage

windspeed  decimal       the speed that air is moving in unknown units

casual     integer       the number of non-registered bike shares for the hour

registered integer       the number of registered bike shares for the hour

count      integer       the total number of bike shares for the hour
--------------------------------------------------------------------

### Comments ###

There were no inherent character columns and the data, where appropriate, was 
already converted to factor-like integer values. As such, in order to better 
map the data set to this exercise, I am adding a `seasonName` column and 
reverting the `season` column into this new column to begin with. Additionally,
for analysis purposes the datetime field would be better broken up into 
components including a simple integer hour of day, individual month value, 
day of week, segment of the day, etc. Some of these transformations are applied
in the code segment that follows.

```{r}
# The path to the local dataset file
csv_file <- file.path(projRoot, "Week5", "BikeSharingDemand.csv")

# Load the data into a data.frame
csv <- read.table(csv_file, header=TRUE, sep=",")

# Revert some columns to character data.
bikes <- data.frame(csv, 
                    seasonName=NA, hourOfDay=NA, 
                    dayOfWeek=NA, dayOfWeekInt=NA, 
                    monthOfYear=NA, segmentOfDay=NA)
bikes[bikes$season == 1,]$seasonName <- "spring"
bikes[bikes$season == 2,]$seasonName <- "summer"
bikes[bikes$season == 3,]$seasonName <- "fall"
bikes[bikes$season == 4,]$seasonName <- "winter"

# 14 Add an integer column for hour of day
bikes$hourOfDay <- lubridate::hour(bikes$datetime)

# 15 Add an factor and integer column for day of week
bikes$dayOfWeek <- as.factor(weekdays(strptime(as.character(bikes$datetime), 
                                               format="%Y-%m-%d %H:%M:%S")))
# 16
bikes$dayOfWeekInt <- as.numeric(bikes$dayOfWeek)

# 17 Add an integer column for month of year
bikes$monthOfYear <- lubridate::month(bikes$datetime)

# 18 Add a character column for segment of day
bikes$segmentOfDay <- ifelse(bikes$hourOfDay >= 5 & bikes$hourOfDay < 12, 
                             1, 
                             ifelse(bikes$hourOfDay >= 12 & bikes$hourOfDay < 17, 
                                    2, 
                                    ifelse(bikes$hourOfDay >= 17 & bikes$hourOfDay < 22, 
                                           3, 4)))
```

### Summary Statistics ###
Using the `summary()` R function, the basic statistics about each attribute are summarized through the raw R output that follows.

```{r}
# Summary
summary(bikes)
```

```{r}
outliers = lofactor()
```

### Correlation Analysis ###

Using the `cor()` R function as shown in the following code, an analysis of correlation between the numeric attributes was performed. 
```{r}

bikeCor <- cor(bikes[, c(12, 6:11,14,16,17,18)])
bikesCorMelt <- reshape2::melt(bikeCor, varnames=c("x", "y"), value.name="Correlation")
bikesCorMelt <- bikesCorMelt[order(bikesCorMelt$Correlation),]

```

```{r, echo=FALSE}
g1 <- ggplot(data=bikesCorMelt, aes(x=x, y=y))
g1 <- g1 + geom_tile(aes(fill=Correlation))
g1 <- g1 + scale_fill_gradient2(low="red", mid="white", high="steelblue", 
                                guide=guide_colorbar(ticks=FALSE, barheight=10),
                                limits=c(-1,1))
g1 <- g1 + theme_minimal()
g1 <- g1 + theme(axis.text.x = element_text(angle=30, vjust=1))
g1 <- g1 + labs(title="Heatmap of Attribute Correlation")
g1

```

As can be seen in the heat map above and values below, humidity is negatively correlated with bike sharing demand across all three measures (casual, registered and count). Likewise, temperature is positively correlated with bike sharing demand in this dataset.

```{r, echo=FALSE}
bikeCor
```

### Entropy Analysis ###

Using Entropy and Information Gain functions developed in a prior exercise, an entropy analysis was performed. Raw entropy of the bike shares per hour was calculated initially.

```{r}
source(file.path(projRoot, "EntropyFunctions.R"), chdir=TRUE)

# Raw Entropy: Casual Bike Sharing
entropy(bikes$casual)

# Raw Entropy: Registered Bike Sharing
entropy(bikes$registered)

# Raw Entropy: Total Bike Sharing
entropy(bikes$count)
```

The `decide()` function from the EntropyFunctions script was used to calculate information gain across all attributes versus the `casual` bike sharing measure. The results were then melted into a long format and sorted for better visualization. The R code is shown below and this process was repeated for the `registered` and `count` measures as well. The `humidify` attribute was found to be the most meaning full, followed by our `hourOfDay` calculated attribute.

```{r}
# Calculate information gain across all categorical and numeric attributes. 
casualEnt <- decide(bikes[,c(2,3,4,5,6,7,8,9,10,14,15,17,18)], 9)

casEntMelt <- reshape2::melt(casualEnt$gains, value.name="info.gain")
casEntMelt <- cbind(casEntMelt, attribute=rownames(casEntMelt))
casEntMelt <- casEntMelt[order(-casEntMelt$info.gain),]
```
```{r, echo=FALSE}
require(ggplot2)
g2 <- ggplot(data=casEntMelt, aes(x=factor(attribute), y=info.gain))
g2 <- g2 + geom_bar(stat="identity")
g2 <- g2 + theme_minimal()
g2 <- g2 + scale_x_discrete(labels=as.character(casEntMelt$attribute), limits=as.character(casEntMelt$attribute))
g2 <- g2 + theme(axis.text.x = element_text(angle=30, vjust=1))
g2 <- g2 + labs(title="Information Gain - Casual Bike Shares by Partition", x="Paritioning Attribute", y="Information Gain")
g2
casEntMelt
```

```{r, echo=FALSE}
regEnt <- decide(bikes[,c(2,3,4,5,6,7,8,9,11,14,15,17,18)], 9)

regEntMelt <- reshape2::melt(regEnt$gains, value.name="info.gain")
regEntMelt <- cbind(regEntMelt, attribute=rownames(regEntMelt))
regEntMelt <- regEntMelt[order(-regEntMelt$info.gain),]
```
```{r, echo=FALSE}
require(ggplot2)
g3 <- ggplot(data=regEntMelt, aes(x=factor(attribute), y=info.gain))
g3 <- g3 + geom_bar(stat="identity")
g3 <- g3 + theme_minimal()
g3 <- g3 + scale_x_discrete(labels=as.character(regEntMelt$attribute), limits=as.character(regEntMelt$attribute))
g3 <- g3 + theme(axis.text.x = element_text(angle=30, vjust=1))
g3 <- g3 + labs(title="Information Gain - Registered User Bike Shares by Partition", x="Paritioning Attribute", y="Information Gain")
g3
regEntMelt
```

```{r, echo=FALSE}
countEnt <- decide(bikes[,c(2,3,4,5,6,7,8,9,12,14,15,17,18)], 9)

countEntMelt <- reshape2::melt(countEnt$gains, value.name="info.gain")
countEntMelt <- cbind(countEntMelt, attribute=rownames(countEntMelt))
countEntMelt <- countEntMelt[order(-countEntMelt$info.gain),]
```
```{r, echo=FALSE}
require(ggplot2)
g4 <- ggplot(data=countEntMelt, aes(x=factor(attribute), y=info.gain))
g4 <- g4 + geom_bar(stat="identity")
g4 <- g4 + theme_minimal()
g4 <- g4 + scale_x_discrete(labels=as.character(countEntMelt$attribute), limits=as.character(countEntMelt$attribute))
g4 <- g4 + theme(axis.text.x = element_text(angle=30, vjust=1))
g4 <- g4 + labs(title="Information Gain - Total Bike Shares by Partition", x="Paritioning Attribute", y="Information Gain")
g4
countEntMelt
```

### References ###

Fanaee-T, Hadi, and Gama, Joao, Event labeling combining ensemble detectors and background knowledge, Progress in Artificial Intelligence (2013): pp. 1-15, Springer Berlin Heidelberg.

Lander, Jared P. "Correlation and Covariance." R for Everyone: Advanced Analytics and Graphics. New York: Addison-Wesley, 2014. N. pag. Print.

