---
title: 'Data Set Profile: Bike Sharing Demand'
author: "Daniel Dittenhafer"
date: "Tuesday, September 27, 2014"
output: pdf_document
---
```{r, echo=FALSE} 
require(ggplot2, quietly=TRUE)
require(lubridate, quietly=TRUE)
projRoot <- "C:/SourceCode/R/DataAcqMgmt"
```
----------------------------------------------------------------------------------------------------
Data Set         Number of              Attribute                        Number of 
Characteristics  Observations Area      Characteristics                  Attributes  Missing Values?
--------------- ------------- --------- -------------------------------- ----------- ---------------
Multivariate    10,886        Business  Categorical, Integer, Date/time  12          No     
--------------------------------------- -------------------------------- ----------- ---------------

### Source ###

[Kaggle.com Competition: Bike Sharing Demand](https://www.kaggle.com/c/bike-sharing-demand): https://www.kaggle.com/c/bike-sharing-demand


### Attribute Information ###

--------------------------------------------------------------------
Field      Data Type     Description
---------  ------------- --------------------------------------------
datetime   date & hour   Min: 01/01/2011 00:00;    
                         Max: 12/19/2012 23:00

season     integer       Categorical;
                         1 = spring;
                         2 = summer;
                         3 = fall;
                         4 = winter;
                         
holiday    boolean       1 = a holiday;
                         0 = not a holiday;

workingday boolean       1 = a work day;
                         0 = weekend or holiday;
                         
weather    integer       Categorical;
                         1) Clear, Few clouds, Partly cloudy, Partly cloudy;
                         2) Mist + Cloudy, Mist + Broken clouds, Mist + Few clouds, Mist; 
                         3) Light Snow, Light Rain + Thunderstorm + Scattered clouds, Light Rain + Scattered clouds;
                         4) Heavy Rain + Ice Pallets + Thunderstorm + Mist, Snow + Fog; 
                         
temp       decimal       temperature in degrees Celsius

atemp      decimal       apparent temperature in degrees Celsius 

humidity   integer       relative humidity percentage

windspeed  decimal       the speed that air is moving in unknown units

casual     integer       the number of non-registered bike shares for the hour

registered integer       the number of registered bike shares for the hour

count      integer       the total number of bike shares for the hour
--------------------------------------------------------------------

### Comments ###

There were no inherent character columns and the data, where appropriate, was 
already converted to factor-like integer values. As such, in order to better 
map the data set to this exercise, I have added a `seasonName` column and 
reverted the `season` column into this new column to begin with. Additionally,
for analysis purposes the datetime field would be better broken up into 
components including a simple integer hour of day, individual month value, 
day of week, segment of the day, etc. Some of these transformations are applied
in the code segment that follows.

```{r}
# The path to the local dataset file
csv_file <- file.path(projRoot, "Week5", "BikeSharingDemand.csv")

# Load the data into a data.frame
csv <- read.table(csv_file, header=TRUE, sep=",")

# Revert season to character data.
bikes <- data.frame(csv, 
                    seasonName=NA, hourOfDay=NA, 
                    dayOfWeek=NA, dayOfWeekInt=NA, 
                    monthOfYear=NA, segmentOfDay=NA)
bikes[bikes$season == 1,]$seasonName <- "spring"
bikes[bikes$season == 2,]$seasonName <- "summer"
bikes[bikes$season == 3,]$seasonName <- "fall"
bikes[bikes$season == 4,]$seasonName <- "winter"

# 14 Add an integer column for hour of day
bikes$hourOfDay <- lubridate::hour(bikes$datetime)

# 15 Add a factor and integer column for day of week
bikes$dayOfWeek <- as.factor(weekdays(strptime(as.character(bikes$datetime), 
                                               format="%Y-%m-%d %H:%M:%S")))
# 16
bikes$dayOfWeekInt <- as.numeric(bikes$dayOfWeek)

# 17 Add an integer column for month of year
bikes$monthOfYear <- lubridate::month(bikes$datetime)

# 18 Add an integer column for segment of day
bikes$segmentOfDay <- ifelse(bikes$hourOfDay >= 5 & bikes$hourOfDay < 12, 
                             1, 
                             ifelse(bikes$hourOfDay >= 12 & bikes$hourOfDay < 17, 
                                    2, 
                                    ifelse(bikes$hourOfDay >= 17 & bikes$hourOfDay < 22, 
                                           3, 4)))
```

### Summary Statistics ###
Using the `summary()` R function, the basic statistics about each attribute are
summarized through the raw R output that follows.

```{r}
# Summary
summary(bikes)
```

### Outlier Analysis ###
Using the Data Mining with R (DMwR) package's `lofactor()` function, an outlier
analysis was performed. Although a complete analysis was performed across all 
categorical and numeric attributes, generally no extreme outliers were detected, 
with a notable exception. The following R code illustrates the approach used in
the outlier analysis with the top 5 outlying points highlighted in the following
charts via a red plus symbol.

```{r}
outlier.scores <- DMwR::lofactor(bikes[,c(5:9, 12)], k=5)

outliers <- order(outlier.scores, decreasing=T)[1:5]

```

One might reasonably expect temperature and apparent temperature to follow one
another more or less linearly. For each unit increase in temperature, apparent
temperature would increase by approximately one unit. As shown in the following 
chart, this is mostly true, except for the August 17, 2012 values. On this day, 
the connection between temperature and apparent temperature appear broken with 
`atemp` stuck at 12.12 $^\circ$C. Without further knowledge of the data's 
origins, one can only speculate as to why this is the case.

```{r, echo=FALSE}
n <- nrow(bikes)
pch <- rep(".", n)
pch[outliers] <- "+"
col <- rep("black", n)
col[outliers] <- "red"

# Plot the outliers along with the other data
pairs(bikes[,c(6:7,12)], pch=pch, col=col, main="Outliers in Context")

# Show the raw data
bikes[lubridate::day(bikes$datetime) == 17 & lubridate::month(bikes$datetime) == 8 & lubridate::year(bikes$datetime) == 2012,c(1, 6:7)]
```

### Correlation Analysis ###

Using R's `cor()` function, as shown in the following code, an analysis of 
correlation between the numeric attributes was performed. 
```{r}

bikeCor <- cor(bikes[, c(12, 6:11,14,16,17,18)])
bikesCorMelt <- reshape2::melt(bikeCor, varnames=c("x", "y"), value.name="Correlation")
bikesCorMelt <- bikesCorMelt[order(bikesCorMelt$Correlation),]

```

```{r, echo=FALSE}
g1 <- ggplot(data=bikesCorMelt, aes(x=x, y=y))
g1 <- g1 + geom_tile(aes(fill=Correlation))
g1 <- g1 + scale_fill_gradient2(low="red", mid="white", high="steelblue", 
                                guide=guide_colorbar(ticks=FALSE, barheight=10),
                                limits=c(-1,1))
g1 <- g1 + theme_minimal()
g1 <- g1 + theme(axis.text.x = element_text(angle=30, vjust=1))
g1 <- g1 + labs(title="Heatmap of Attribute Correlation")
g1

```

As can be seen in the heat map above and values below, humidity is negatively
correlated with bike sharing demand across all three measures (casual, 
registered and count). Likewise, temperature is positively correlated with bike
sharing demand in this data set.

```{r, echo=FALSE}
bikeCor
```

### Entropy Analysis ###

Using Entropy and Information Gain functions developed in a prior exercise, 
an entropy analysis was performed. Raw entropy of the bike shares per hour was
calculated initially.

```{r}
source(file.path(projRoot, "EntropyFunctions.R"), chdir=TRUE)

# Raw Entropy: Casual Bike Sharing
entropy(bikes$casual)

# Raw Entropy: Registered Bike Sharing
entropy(bikes$registered)

# Raw Entropy: Total Bike Sharing
entropy(bikes$count)
```

The `decide()` function from the EntropyFunctions script was used to calculate 
information gain across all attributes versus the `casual` bike sharing measure. 
The results were then melted into a long format and sorted for better visualization. 
The R code is shown below and this process was repeated for the `registered` and 
`count` measures as well. None of the attributes produced particularly staggering
information gain, but the `humidity` attribute was found to be the most meaningful, 
followed by the `hourOfDay` calculated attribute and `temp`/`atemp` attributes.

```{r}
# Calculate information gain across all categorical and numeric attributes. 
casualEnt <- decide(bikes[,c(2,3,4,5,6,7,8,9,10,14,15,17,18)], 9)

casEntMelt <- reshape2::melt(casualEnt$gains, value.name="info.gain")
casEntMelt <- cbind(casEntMelt, attribute=rownames(casEntMelt))
casEntMelt <- casEntMelt[order(-casEntMelt$info.gain),]
```
```{r, echo=FALSE}
require(ggplot2)
g2 <- ggplot(data=casEntMelt, aes(x=factor(attribute), y=info.gain))
g2 <- g2 + geom_bar(stat="identity")
g2 <- g2 + theme_minimal()
g2 <- g2 + scale_x_discrete(labels=as.character(casEntMelt$attribute), limits=as.character(casEntMelt$attribute))
g2 <- g2 + theme(axis.text.x = element_text(angle=30, vjust=1))
g2 <- g2 + labs(title="Information Gain - Casual Bike Shares by Partition", x="Paritioning Attribute", y="Information Gain")
g2
casEntMelt
```

```{r, echo=FALSE}
regEnt <- decide(bikes[,c(2,3,4,5,6,7,8,9,11,14,15,17,18)], 9)

regEntMelt <- reshape2::melt(regEnt$gains, value.name="info.gain")
regEntMelt <- cbind(regEntMelt, attribute=rownames(regEntMelt))
regEntMelt <- regEntMelt[order(-regEntMelt$info.gain),]
```
```{r, echo=FALSE}
require(ggplot2)
g3 <- ggplot(data=regEntMelt, aes(x=factor(attribute), y=info.gain))
g3 <- g3 + geom_bar(stat="identity")
g3 <- g3 + theme_minimal()
g3 <- g3 + scale_x_discrete(labels=as.character(regEntMelt$attribute), limits=as.character(regEntMelt$attribute))
g3 <- g3 + theme(axis.text.x = element_text(angle=30, vjust=1))
g3 <- g3 + labs(title="Information Gain - Registered User Bike Shares by Partition", x="Paritioning Attribute", y="Information Gain")
g3
regEntMelt
```

```{r, echo=FALSE}
countEnt <- decide(bikes[,c(2,3,4,5,6,7,8,9,12,14,15,17,18)], 9)

countEntMelt <- reshape2::melt(countEnt$gains, value.name="info.gain")
countEntMelt <- cbind(countEntMelt, attribute=rownames(countEntMelt))
countEntMelt <- countEntMelt[order(-countEntMelt$info.gain),]
```
```{r, echo=FALSE}
require(ggplot2)
g4 <- ggplot(data=countEntMelt, aes(x=factor(attribute), y=info.gain))
g4 <- g4 + geom_bar(stat="identity")
g4 <- g4 + theme_minimal()
g4 <- g4 + scale_x_discrete(labels=as.character(countEntMelt$attribute), limits=as.character(countEntMelt$attribute))
g4 <- g4 + theme(axis.text.x = element_text(angle=30, vjust=1))
g4 <- g4 + labs(title="Information Gain - Total Bike Shares by Partition", x="Paritioning Attribute", y="Information Gain")
g4
countEntMelt
```

### Source Code ###
The raw R markdown code used to produce this data set profile can be found 
[on GitHub, in my DataAcqMgmt repository](https://raw.githubusercontent.com/dwdii/DataAcqMgmt/master/Week5/Project2-ProfilingADataSet-Bsd.Rmd).

### References ###

Fanaee-T, Hadi, and Gama, Joao, Event labeling combining ensemble detectors and background knowledge, Progress in Artificial Intelligence (2013): pp. 1-15, Springer Berlin Heidelberg.

Lander, Jared P. "Correlation and Covariance." R for Everyone: Advanced Analytics and Graphics. New York: Addison-Wesley, 2014. N. pag. Print.

Zhao, Yanchang. "[Outlier Detection - RDataMining.com: R and Data Mining](http://www.rdatamining.com/examples/outlier-detection)" RDataMining.com: R and Data Mining. N.p., 2014. Web. 27 Sept. 2014.
